---
title: "Biostat M280 Homework 3"
subtitle: Due Mar 2 @ 11:59PM
output: html_document
---



## Q1 LA City Employee Payroll

The `/home/m280-data/la_payroll/LA_City_Employee_Payroll.csv` file on teaching server contains payroll information of LA City employees in years 2013-2017. It was downloaded from [LA City Controller's Office](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym). Make a Shiny app to facilitate exploratory data analysis. 

1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.


```{r}
library("tidyverse")
```

```{r}

#Read in RDS file
la_payroll <- read_csv("/home/m280-data/la_payroll/LA_City_Employee_Payroll.csv")

saveRDS(la_payroll, "la_payroll.rds", compress = TRUE)
LApayroll <- readRDS("la_payroll.rds")


#Rename the names of the columns
names(LApayroll) <- str_replace_all(names(LApayroll), " ", "_")

data_la_payroll <- 
  LApayroll %>% 
  select(Row_ID, Base_Pay, Overtime_Pay, Total_Payments, `Other_Pay_(Payroll_Explorer)`, Department_Title, Job_Class_Title, Year)  
  

#remove dollar signs
data_la_payroll$Base_Pay <- as.numeric(gsub("\\$", "", data_la_payroll$Base_Pay))
data_la_payroll$Total_Payments <- as.numeric(gsub("\\$", "", data_la_payroll$Total_Payments))
data_la_payroll$Overtime_Pay <- as.numeric(gsub("\\$", "", data_la_payroll$Overtime_Pay))
data_la_payroll$`Other_Pay_(Payroll_Explorer)` <- as.numeric(gsub("\\$", "", data_la_payroll$`Other_Pay_(Payroll_Explorer)`))

```

0. **Total payroll by LA City**. Visualize the total LA City payroll of each year, with breakdown into base pay, overtime pay, and other pay.

0. **Who earned most?** Visualize the payroll information (total payment with breakdown into base pay, overtime pay, and other pay, Department, Job Title) of the top $n$ highest paid LA City employees in a specific year. User specifies $n$ (default 10) and year (default 2017).

0. **Which departments earn most?** Visualize the mean or median payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ earning departments. User specifies $n$ (default 5), year (default 2017), and method (mean or median, default median).

0. **Which departments cost most?** Visualize the total payroll, with breakdown into base pay, overtime pay, and other pay, of top $n$ expensive departments. User specifies $n$ (default 5) and year (default 2017).

0. Visualize any other information you are interested in.

0. Publish your Shiny app to <https://www.shinyapps.io> and share the link.

## Q2 LA City Parking War

The SQLite database `/home/m280-data/la_parking/LA_Parking_Citations.sqlite` on teaching server contains information about parking tickets in LA City. It was downloaded from [LA Open Data Portal](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np). Connect to the database and answer following questions using plots and summary statistics. In this exercise, you are **not** allowed to load whole data into memory. Use the _transform in database, plot in R_ strategy.


```{r}
library("DBI")
library("RSQLite")
library(anytime)
```


```{r}

if (Sys.info()[["sysname"]] == "Linux") {
  db <- dbConnect(RSQLite::SQLite(),
                  dbname = "/home/m280-data/la_parking/LA_Parking_Citations.sqlite")
} else if (Sys.info()[["sysname"]] == "Darwin") {
  db <- dbConnect(RSQLite::SQLite(),
                  dbname = "./LA_Parking_Citations.sqlite")
}

dbListTables(db)
 
latix_sql <- dplyr::tbl(db, "latix")



db1 <- dbConnect(RSQLite::SQLite(),
                  dbname = "/home/m280-data/la_parking/LA_Parking_Citations_Extra.sqlite")
dbListTables(db1)

new_db <- dbSendQuery(db1, "Select Issue_Year, Issue_Month, Issue_Day, Issue_Hour, Issue_Minute, Make, Violation_Description FROM latix")


latix_sql1 <- dplyr::tbl(new_db, "latix")




```


1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?

How many tickets (how many rows)?

```{r}

latix_sql <- dplyr::tbl(db, "latix")
str(latix_sql)

count(latix_sql)

```


  *There are 4,044,488 tickets in this data set.*

**Which time period do these tickets span?**
```{r}
latix <- as.data.frame(latix_sql)

latix$Issue_DateTime <- anydate(latix$Issue_DateTime)

sorted_dates <- sort(latix$Issue_DateTime)
head(sorted_dates)
tail(sorted_dates)



```

*The time period spans from 4-27-2010 to 12-29-2017*



**Which years have most data?**
```{r}
years <- substr(latix$Issue_DateTime, 1, 4)

table(years)

```
*The year 2015 has the most data*



0. When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

```{r}


```

0. Which car makes received most citations?
```{r}
model <- latix$Make

table(model)
max(table(model))


```

*TOYT has the most citations*
0. How many different colors of cars were ticketed? Which color attracted most tickets?
```{r}

color <- latix$Color
table(color)
max(table(color))

```
*There are 65 different colors. BK is the color that was most ticketed*

0. What are the most common ticket types?

0. How much money was collected on parking tickets in 2015 and 2016?

0. Visualize any other information you are interested in.